---
title: "Projecting actor locations and modeling dyadic interactions with PALS"
output: rmarkdown::html_vignette
bibliography: palsr.bib
vignette: >
  %\VignetteIndexEntry{Projecting actor locations and modeling dyadic interactions with PALS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 96
)
set.seed(1)
```

## Overview

Many actors in the social world --- armed groups, firms, diplomats, migrating
populations --- have no fixed location. They move through space over time, and
*where* two such actors interact is itself an outcome worth modeling. The
**Projected Actor Location (PALS)** method [@kim2023pals] addresses this by
projecting where a mobile actor "is" at any moment from the spatiotemporal
history of its past interactions, using exponential-smoothing weights that
favour recent and nearby events.

The `palsr` package implements the full workflow:

1. build a validated table of dyadic events (`pal_events()`);
2. estimate the smoothing parameters by minimizing great-circle prediction
   error (`estimate_pals()`);
3. project actor locations at arbitrary times (`project_pals()`);
4. predict dyadic interaction locations and build distance covariates
   (`predict_event_locations()`, `pal_distance()`);
5. quantify uncertainty with a nonparametric bootstrap and pool with Rubin's
   Rules (`bootstrap_pals()`, `pool_rubin()`).

```{r load}
library(palsr)
```

## The method in brief

For a focal actor $i$ at prediction time $t$, PALS forms a recency-weighted
mean of the locations of $i$'s own past events (the *focal* component) and a
recency-weighted mean of the locations of events involving $i$'s past
interaction partners, or *alters* (the *alter* component). The projected
location is a convex combination of the two,
$$
g_i(t) = (1-\pi)\sum_e W_i(e)\, g(e) \;+\; \pi \sum_e W_k(e)\, g(e),
\qquad \pi = \mathrm{logistic}(\gamma + \eta\, v),
$$
where the weights decay with event age --- $W_i(e) \propto (\text{age}^{\alpha})^{-1}$
for the focal actor and analogously with $\beta$ for the alters --- and the
mixing weight $\pi$ depends on how active the focal actor is relative to its
alters through the event-count ratio $v$. The four parameters are therefore:

| Parameter | Role |
|-----------|------|
| $\alpha$  | decay of the focal actor's own history |
| $\beta$   | decay of the alters' histories |
| $\gamma$  | intercept of the focal-vs-alter mixing weight |
| $\eta$    | dependence of the mixing weight on relative activity |

A reduced **one-parameter** model fixes $\pi = 0$ (focal history only) and
estimates $\alpha$ alone; it is fast and surprisingly competitive.

## A worked example

The package ships a deterministic simulated dataset, `nigeria_sim`, of 1,500
dyadic conflict events among 25 mobile actors between 2000 and 2016. (The
original application uses ACLED data, which are not redistributable.)

```{r data}
data(nigeria_sim)
nigeria_sim
summary(nigeria_sim)
```

You can build your own `pal_events` object from any data frame by naming the
actor, time, longitude and latitude columns:

```{r build}
raw <- data.frame(
  from = c("A", "A", "B"),
  to   = c("B", "C", "C"),
  when = as.Date(c("2001-01-01", "2001-06-01", "2002-01-01")),
  x    = c(7.1, 8.0, 7.5),
  y    = c(9.0, 9.4, 10.1)
)
pal_events(raw, actor1 = "from", actor2 = "to",
           time = "when", lon = "x", lat = "y")
```

### Estimating the parameters

Estimation marches forward through time: every event is predicted using only
events strictly earlier than it, and the parameters minimize the mean
great-circle (Haversine) distance between predicted and observed locations.

```{r fit-one}
fit1 <- estimate_pals(nigeria_sim, model = "one")
fit1
coef(fit1)
```

The full four-parameter model adds the alter component. We cap the optimizer
iterations here purely to keep the vignette quick:

```{r fit-four}
fit4 <- estimate_pals(nigeria_sim, model = "four",
                      control = list(maxit = 60))
coef(fit4)
```

### Projecting actor locations

With a fitted model (or a hand-specified `pals_params()`), project where each
actor is at a given time:

```{r project}
pal_2015 <- project_pals(nigeria_sim,
                         predict_time = as.Date("2015-01-01"),
                         params = fit1)
head(pal_2015)
```

```{r map, fig.alt = "Projected actor locations on 2015-01-01"}
library(ggplot2)
ggplot(pal_2015, aes(lon, lat)) +
  geom_point(colour = "#2b6cb0", size = 2) +
  geom_text(aes(label = actor), vjust = -0.8, size = 3) +
  labs(title = "Projected actor locations, 2015-01-01",
       x = "Longitude", y = "Latitude") +
  theme_minimal()
```

### Predicting interaction locations and dyadic distances

The predicted location of an interaction between two actors is the mean of
their two projected locations. Supplying observed coordinates scores the
prediction in kilometres:

```{r predict-events}
targets <- nigeria_sim[nigeria_sim$time > as.Date("2014-01-01"), ]
scored  <- predict_event_locations(nigeria_sim, targets, fit1)
summary(scored$error_km)
```

The dyadic distance between two actors' projected locations is the key
covariate for modeling who interacts with whom:

```{r distance}
dyads <- data.frame(actor1 = "G01", actor2 = "G02",
                    time = as.Date("2014-06-01"))
pal_distance(nigeria_sim, dyads, fit1, transform = "log")
```

## Uncertainty: bootstrap and Rubin's Rules

`bootstrap_pals()` resamples events with replacement and re-estimates the model
on each replicate, yielding bootstrap standard errors and percentile intervals.
(We use a small number of replicates here for speed; the paper uses ten.)

```{r bootstrap}
bt <- bootstrap_pals(nigeria_sim, R = 10, model = "one", seed = 1)
summary(bt)
```

When a downstream estimand (say, a regression coefficient using PAL distances)
is computed on each replicate, treat the replicates as multiple imputations and
combine them with Rubin's Rules, which propagate both within- and
between-replicate uncertainty:

```{r rubin}
q <- c(1.10, 0.95, 1.20, 1.05, 0.98)   # per-replicate estimates
u <- c(0.04, 0.05, 0.045, 0.038, 0.052) # per-replicate variances
pool_rubin(q, u, df = TRUE, dfcom = 100)
```

## References
